---
title: "How to Learn R"
author: "Sarah Manski"
date: "April 2023"
output: html_document
---
With updates by Katherine Muller, October 2025

```{r setup, include=FALSE}
# This is the setup chunk that includes some settings for your document
# Here "include = FALSE" means that you won't see this chunk when you knit your document

# The command below sets a global default option for the rest of the chunks where "echo = TRUE" so you will see both the code and results output, unless otherwise specified in the individual chunk header
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise 1: Data manipulation and summary with dplyr

In this exercise, you are given an example of data manipulation using the *tidyverse* package *dplyr*.  In part (a), you use strategic searching tips to find out what the code does and write out the steps in the code comments.  In part (b), you use the code from part (a) to solve a similar data manipulation task.


### 1a) Learning by example

```{r ex1a}
# Step 1:
library(agridat)
library(tidyverse)

# Step 2:
data(ilri.sheep)

# Step 3:
sheep <- ilri.sheep

# Step 4:
sheep94 <- sheep %>% filter(year == 94, !is.na(weanwt)) 

# Step 5:
sheep94 <- sheep94 %>%
  mutate(wtchange = weanwt - birthwt)

# Step 6:
sheepSummary <- sheep94 %>% group_by(gen) %>% summarize(mean_birthwt = mean(birthwt),
      mean_weanwt = mean(weanwt), mean_wtchange = mean(wtchange), n= n())

# Step 7:
head(sheep94)

# Step 8:
rm(sheep, sheep94, ilri.sheep)
```


### 1b) Adjusting examples to your needs
Suppose now you are given the following task, which requires similar data manipulation as the example in part (a).


Task: Using the functions from the *tidyverse* package *dplyr* and the dataset ilri.sheep from the *agridat* package, complete the following. For all male lambs with a recorded wean age, calculate the standard deviation and median wean age and ewe (dam) age and the maximum ewe (dam) age for each year and ewe genotype.  


Use the sample code from 1a and the steps written in the code chunk below to complete the above task. 

```{r ex1b}
# Step 1: Load required packages

# Step 2: Load required data and view its documentation

# Step 3: Copy the ilri.sheep data into a new object called sheep

# Step 4: Filter data to only include male lambs with a recorded wean age

# Step 5: Group data by year and ewe genotype

# Step 6: Summarize data by calculating standard deviation and median for both wean age and dam age as well as maximum dam age

# Step 7: Inspect the summarized data

# Step 8: Clear one or more objects from your working environment

```



## Exercise 2: Plotting with ggplot2

In this exercise, you are given an example of plot creation using the *tidyverse* package *ggplot2*.  In part (a), you use strategic searching tips to find out what the code does and write out the steps in the code comments.  In part (b), you use the code from part (a) to solve a similar data plotting task.


### 2a) Learning by example



```{r ex2a, warning = FALSE}
# Step 1: 
library(agridat)
library(tidyverse)

# Step 2: 
data(ilri.sheep)
sheep <- ilri.sheep

# Step 3: 
ggplot(data = sheep) + 
  # Step 4: 
  geom_point(aes(x = weanage, y=weanwt, color = gen)) + 
  # Step 5:
  labs(x = "Lamb weaning age", y = "Lamb weaning weight", title = "Lamb weaning age and weight by genotype") +
  # Step 6:
  facet_wrap(~year)
```

### 2b) Adjusting examples to your needs

Suppose now you are given the following task, which requires similar data plotting as the example in part (a).


Task: Use the functions from the *tidyverse* package *ggplot2* and the dataset ilri.sheep from the *agridat* package. Create scatter plots of birth weight vs wean weight grouped by lamb genotype, with appropriate axis labels and title, and where the shape of the points corresponds to the lamb sex and the opacity (alpha) of the points corresponds to lamb wean age.


Use the sample code from 2a and the steps written in the code chunk below to complete the above task. 



```{r ex2b, warning = FALSE}
# Step 1: Load required packages


# Step 2: Load required data and save


# Step 3: Call ggplot with sheep data

  # Step 4: Create a scatter plot with birth weight for x axis, wean weight for y axis, shape for sex and alpha for wean age

  # Step 5: Include descriptive labels

  # Step 6: Split plot by lamb genotype

```



